---
title: "HyperCom Overview"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(HyperCom)
```

```{r}
library(dplyr)
library(ggplot2)
```

## Load in example data
HyperCom requires three inputs:

1. A dense gene x cell matrix with genes as row names and cells as column names

2. A dataframe of cell metadata containing sample and optionally group columns which need to be specified as variables

3. A dataframe of ligand-receptor (LR) interactions. HyperCom contains 2 built in LR databases formatted properly:

    + CellChat human `cellchat.hs.lrdb`
  
    + CellChat mouse `cellchat.mm.lrdb`

In this vignette we will use a 1000 cell subset of heart failure from [Alexanian et al, 2021](https://doi.org/10.1038/s41586-021-03674-1).
This example subset is included in the package as `alexanian2021.counts` and `alexanian2021.metadata`.
```{r}
counts <- alexanian2021.counts
dim(counts)
head(rownames(counts))
head(colnames(counts))
```

```{r}
meta.data <- alexanian2021.metadata
sample.column <- "sample"
group.column <- "condition"
head(meta.data)
```

```{r}
ggplot(meta.data, aes(x=UMAP_1, y=UMAP_2, colour=celltype)) +
  geom_point(size=0.5) +
  theme_classic()
```

For the LR interactions we will use the CellChat murine database.
```{r}
lrdb <- cellchat.mm.lrdb
dim(lrdb)
```

```{r}
head(lrdb)
```

## Running HyperCom
The standard pipeline can be run using the `HyperCom` function. This vignette details each component function of the pipeline.
```{r, eval=FALSE}
HyperCom(lrdb=cellchat.mm.lrdb, counts=alexanian2021.counts, metadata=alexanian2021.metadata, sample="sample",
         group="condition", keep=NULL, significance=FALSE, parallel=FALSE, cl=NULL)
```

### Preprocessing
In order to reduce memory usage, we first filter the counts matrix to only genes present in the LR interactions with expression greater than 0.
The function also transposes the matrix.
```{r}
lr.counts <- filter_counts(counts=counts, lrdb=lrdb, outfile=NULL)
dim(lr.counts)
```

Next we filter the ligand receptor interactions to only interactions where all genes are present in the filtered expression matrix.
```{r}
lrs <- generate_lrs_table(counts=lr.counts, lrdb=lrdb, outfile=NULL)
dim(lrs)
```

### Generating Hypergraph
To generate the hyperedges, which are edges that contain multiple nodes, we use the expression matrix LR interaction to create 3 types of hyperedges:

1. expression edges - edges containing cells that express a given ligand or receptor

2. complex edges - edges containing the genes that comprise a receptor complex

3. interaction edges - edges containing the genes that comprise an LR interaction

HyperCom represents these hyperedges as a dataframe with an edge and node column where the node belongs to the edge. The weight is set to 1 for each edge. 
```{r}
hyperedges <- generate_hyperedges(counts=lr.counts, lrs=lrs, outfile=NULL)
head(hyperedges)
```

We can represent this hypergraph as an adjacency matrix where the nodes are the cells, ligand genes, and receptor genes.
Each entry is the number hyperedges shared by nodes.
```{r}
adj <- generate_adjacency_matrix(hyperedges=hyperedges, outfile=NULL)
dim(adj)
```

### Calculating Network Connectivity
In order to calculate the network connectivity of the hypergraph we perform a random walk on the adjacency matrix to produce an influence matrix.
```{r}
infMat <- generate_infMat(adj=adj, outfile=NULL)
dim(infMat)
```

## Downstream Analysis
To account for genes now being present in the matrices, we then process the metadata to contain information on the genes as well as cell metadata.
Additionally to reduce memory, we reduce the number of columns to just node, sample, and group. For this vignettes' visualization, we will also keep UMAP_1 and UMAP_2
```{r}
metadata <- generate_metadata(infMat=infMat, metadata=meta.data, sample=sample.column, group=group.column,
                              keep=c("UMAP_1", "UMAP_2"), outfile=NULL)
head(metadata)
```

### Scoring Network Activity
We can score network activity using the influence matrix to understand globally if cells are sending or receiving signals.
HyperCom scoring generates a sender and receiver score based on a given matrix and uses both to create a composite HyperCom score. This HyperCom score ranges from -1 to 1 going from better senders to better receivers.
```{r}
network.score <- score_hypercom(infMat=infMat, metadata=metadata)
head(network.score)
```

Here we can see that globally, communication is higher in TAC.
```{r}
network.score <- left_join(network.score, metadata, by=join_by(cell))

transparent <- rgb(255, 255, 255, max=255, alpha = 0)
ggplot(network.score, aes(x=as.numeric(UMAP_1), y=as.numeric(UMAP_2), colour=hypercom)) +
  geom_point(size=0.5) +
  scale_color_gradient2(low="blue", mid=transparent, high="red", limits=c(-1,1)) +
  facet_wrap(vars(group)) +
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme_classic()
```

### Identifying Important LR Pairs
We can also identify important LR pairs by using HyperCom scoring on LR pairs.
This prioritization scores LR pairs to find highly communicative LR pairs that are optionally group specific.
A p-value can be calculated for each LR pair, but doing so will vastly increase the time to run. By default the `significance` arg is FALSE.
The `prioritize_lr` function can be run sequentially or parallelly and by default the `parallel` arg is FALSE.
The parallel implementation uses the `parallelly` package, but the `cl` can be supplied any object of class `cluster` depending on your system's architecture.
```{r}
priority <- prioritize_lr(lrs=lrs, adj=adj, infMat=infMat, metadata=metadata, significance=FALSE, parallel=TRUE, cl=NULL,
                          outfile=NULL)
head(priority)
```

We see that Thbs1 Sdc4 is the highest scoring LR pair. We can calculate HyperCom scores for this pair.
As a note, the `score_hypercom` function can also take a vector of receptor genes if the receptor is a complex.
```{r}
Thbs1_Sdc4 <- score_hypercom(infMat=infMat, metadata=metadata, ligand="Thbs1", receptors="Sdc4")
head(Thbs1_Sdc4)
```

If we plot the sender and receiver score we see both higher densities of sender and receiver cells in TAC.
```{r}
Thbs1_Sdc4 <- left_join(Thbs1_Sdc4, metadata, by=join_by(cell))
ggplot(Thbs1_Sdc4, aes(x=sender, y=receiver, colour=group)) +
  geom_density_2d() +
  theme_classic()
```

### Side Note: Communication Clusters
Although clustering is not necessary for any step in HyperCom, it may be useful to conceptualize how groups of cells are communicating.
We can cluster the influence matrix to generate communication clusters. Here we use Seurat's implementation of Louvain clustering.
```{r}
library(Seurat)
generate_com_clusters <- function(infMat, metadata){
  com.seurat <- CreateSeuratObject(t(infMat), meta.data = metadata, project="communication")
  com.seurat <- NormalizeData(com.seurat, verbose=FALSE)
  com.seurat <- FindVariableFeatures(com.seurat, selection.method = "vst", nfeatures = 2000, verbose=FALSE)
  all.genes <- rownames(com.seurat)
  com.seurat <- ScaleData(com.seurat, features = all.genes, verbose=FALSE)
  com.seurat <- RunPCA(com.seurat, features = VariableFeatures(object = com.seurat), verbose=FALSE)
  com.seurat <- FindNeighbors(com.seurat, dims = 1:10, verbose=FALSE)
  com.seurat <- FindClusters(com.seurat, resolution = 0.5, cluster.name = "com_clusters", verbose=FALSE)
  com.seurat <- RunUMAP(com.seurat, dims = 1:10, verbose=FALSE)
  
  hyper.com_clusters <- dplyr::select(com.seurat@meta.data, cell, com_clusters)
  metadata <- left_join(metadata, hyper.com_clusters, by=join_by(cell))
  metadata$com_clusters <- factor(metadata$com_clusters)
  
  return(metadata)
}
```

We can see that communication clusters do not necessarily line up with the annotated cell types.
```{r}
com_clusters <- generate_com_clusters(infMat=infMat, metadata=metadata)

annotations <- select(meta.data, celltype)
annotations$cell <- rownames(annotations)
com_clusters <- left_join(com_clusters, annotations, by=join_by(cell))
com_clusters <- filter(com_clusters, group != "gene")

ggplot(com_clusters, aes(x=com_clusters, fill=celltype)) +
  geom_bar() +
  theme_classic()
```
